---
title: "Data Visualization techniques using R (ggplot2)"
author: "Vibs"
date: "March 15, 2018"
output:
  html_document: default
  pdf_document: default
---

# This is a demonstration of different visualization techniques using R (ggplot2 package)

We will use 2 datasets for this purpose:

* iris
* diamonds

I have used the below video from edureka as an inspiration but tried the plots on a different dataset to just try it out with a little bit of customization from my side.

Edureka Tutorial link 

* [ggplot2 Tutorial | ggplot2 In R Tutorial | Data Visualization In R | R Training | Edureka](https://www.youtube.com/watch?v=N5gYo43oLE8)

Let us start with the basic graphs using iris dataset

```{r}

library(ggplot2)

ggplot(iris, aes(y=Petal.Length, x= Petal.Width)) + geom_point()

```

Add color to the aesthetics. it can be added on categorical variable and legend will be auto populated

```{r}

ggplot(iris, aes(y=Petal.Length, x= Petal.Width, col=Species)) + geom_point()

```

Add shape to the aesthetics. it can be added on categorical variable and legend will be auto populated

```{r}

ggplot(iris, aes(y=Petal.Length, x= Petal.Width, col=Species, shape=Species)) + geom_point()

```

Let us move to diamonds dataset to move to the intermediate level of plots

## Histogram

Histogram is used for visualizing continuous variable distribution.

```{r}

library(dplyr)

str(diamonds)

```

Add the number of bins or number of histograms needed

```{r}

ggplot(diamonds, aes(x=price)) +geom_histogram(bins=50)

```

Add color using fill and col to give boundary colors

```{r}
ggplot(diamonds, aes(x=price)) +geom_histogram(bins=50, fill='palegreen4', col='red')

```

Use fill as an aesthetic (here color is a column name in daiamonds dateset). fill can be used on categorical variables

```{r}
ggplot(diamonds, aes(x=price,fill=color)) +geom_histogram(bins=50)

```

If we use fill as an attribute in geom_histogram it will override the aesthtic values

```{r}
ggplot(diamonds, aes(x=price,fill=color)) +geom_histogram(bins=50,fill='palegreen4', col='red')

```

If we use position ='fill' in the geom_histogram function then we get the proportion instead of the count

```{r}
ggplot(diamonds, aes(x=price,fill=color)) +geom_histogram(bins=50, position='fill')

```

## Bar Plot

Bar plot is used for visualizing categorical variable distribution

Put a categorical variable in aesthetic
```{r}
ggplot(diamonds, aes(x=cut)) +geom_bar()

```

Use fill as an aesthetic (here clarity is a column name in daiamonds dateset). fill can be used on categorical variables
Thus, we can visualize 2 categorical variables in  a single graph.

```{r}
ggplot(diamonds, aes(x=cut, fill=clarity)) +geom_bar(position = 'fill')

```

## Frequency Polygon

Frequency Polygon as an alternative to histogram for continuous variable distribution

```{r}
ggplot(diamonds, aes(x=price)) +geom_freqpoly(bins = 50)

```

Play with the size of the lines

```{r}
ggplot(diamonds, aes(x=price)) +geom_freqpoly(bins = 50, size=2)

```

Add multiple frequency lines based on a catgorical variable using col in the aesthetics

```{r}
ggplot(diamonds, aes(x=price, col=cut)) +geom_freqpoly(bins = 50, size=1)


```

## Box Plots

Box Plots to understand how does a continuous variable change w.r.t a categorical variable

```{r}
ggplot(diamonds, aes(x=factor(carat), y=price)) +geom_boxplot()

```

How does price (continuous variable) change w.r.t. cut (categorical variable) of the diamonds

```{r}
ggplot(diamonds, aes(x=cut, y=price)) +geom_boxplot()

```

multi-variate analysis: add color to use 1 more categorical variable. use fill in the aesthetic on the color column

```{r}
ggplot(diamonds, aes(x=cut, y=price, fill=color)) +geom_boxplot()

```

## Smooth line

Smooth line for continuous variable to continuous variable analysis. grey area shown is the error.

```{r}
ggplot(diamonds, aes(x=carat, y=price)) +geom_smooth()

```

We can remove the error by using se=FALSE
```{r}
ggplot(diamonds, aes(x=carat, y=price)) +geom_smooth(se=FALSE)

```

Add multiple lines using a categorical variable with the color aesthetic

```{r}
ggplot(diamonds, aes(x=carat, y=price, color=cut)) +geom_smooth(se=FALSE)

```

method = lm means linear model
Use both point and smooth geometry
```{r}
ggplot(diamonds, aes(x=carat, y=price)) + geom_point() +geom_smooth(method='lm',se=FALSE)

```

And we can add a categorical variable as well by using color

```{r}
ggplot(diamonds, aes(x=carat, y=price, color=cut)) + geom_point() +geom_smooth(method='lm',se=FALSE)

```

Added 1 more categorical variable using shape aesthetic

```{r}
ggplot(diamonds, aes(x=carat, y=price, color=cut, shape=clarity)) + geom_point() +geom_smooth(method='lm',se=FALSE)

```

This one became too messy. Maybe we should try and ignore putting so many variables in a single plot for the ease of understanding the pattern displayed.

## Faceting (to facet the data into groups)

The previous graph was to chaotic, so we can facet the data into groups based on a categorical variable
Now instead of having 5 different colored lines and dots oin a single graph
We will have 5 graphs based on 5 categories of cut column

```{r}
ggplot(diamonds, aes(x=carat, y=price, color=cut)) + geom_point() +geom_smooth(method='lm',se=FALSE) + facet_grid(.~cut)

```

## Theme Layer for presentation purposes

Store the graph in a object and add labels theme

```{r}
obj1 <- ggplot(diamonds, aes(x=cut, y=price, fill=color)) +geom_boxplot()
obj2 <- obj1 + labs(title='My Title',x='my x axis label',y='my y axis label',fill='my legends title')
obj2

```

Add theme layer to give a theme to the plot

```{r}
obj3 <- obj2 + theme(panel.background = element_rect(fill='palegreen4'))
obj3

```

The font of title of the plot can be changed using the below code
hjust to align the title at the center of the plot

```{r}
obj4 <- obj3 + theme(plot.title = element_text(hjust = 0.5, face='bold', colour = 'red'))
obj4

```

Data mapped on y axis is price which is a continuous variable hence we can use scale_y_continuous function to make changes to the scale of y
Similarly if it was categorical we should have used scale_y_discrete
Here we want to scale down the outlier values and put the max of 10000

```{r}
obj4+scale_y_continuous(limits=c(0,10000))

```

Let us try theme on another plot.

```{r}
g1<- ggplot(diamonds, aes(x=carat, y=price, color=cut)) + geom_point() +geom_smooth(method='lm',se=FALSE) + facet_grid(.~cut)
g1

```

```{r}
g2 <- g1+theme(panel.background = element_rect(fill='grey'))
g2

```

Fill the legend with a background
```{r}
g3<- g2 + theme(legend.background = element_rect(fill='grey'))
g3

```

Plot background
```{r}
g3+theme(plot.background = element_rect(fill = 'green'))

```

Thus, we have used almost all basic building blocks of ggplot2 package and are good to go and visualize the world ourselves!

Happy reading,

Vibs!